---
title: "Exploratory Analysis Script"
author: "Vijay Panthayi"
date: "08/20/2023"
output: html_document
---

#Preface

This Quarto file loads the cleaned data and does some exploring.


# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyverse)
library(tidyr)
library(ggrepel)
library(ggtext)
```

First we need to load the cleaned data.

```{r}
#Path to data
data_location1 <- here::here("data","processed_data","carbon_index_data.rds")
data_location2 <- here::here("data","processed_data","carbon_stock_data.rds")
data_location3 <- here::here("data","processed_data","forest_area_data.rds")
data_location4 <- here::here("data","processed_data","forest_index_data.rds")
data_location5 <- here::here("data","processed_data","forest_share_data.rds")
data_location6 <- here::here("data","processed_data","land_area_data.rds")
data_location7 <- here::here("data","processed_data","temp_data_clean.rds")
#load data
carbon_index <- readRDS(data_location1)
carbon_stock <- readRDS(data_location2)
forest_area <- readRDS(data_location3)
forest_index <- readRDS(data_location4)
forest_share <- readRDS(data_location5)
land_area <- readRDS(data_location6)
temperature <- readRDS(data_location7)
```


# Data exploration through tables

Let's start with the climate change temperature data sheet.

Showing a bit of code to produce and save a summary table.

```{r}
summary_temp = skimr::skim(temperature)
print(summary_temp)
# save to file
summarytable_file = here("results", "summary_temp_table.rds")
saveRDS(summary_temp, file = summarytable_file)
```

We can take this time to conjure up a summary table for all the Forest and Carbon data sheets.

```{r}
#carbon_index data
summary_carbon_index = skimr::skim(carbon_index)
print(summary_carbon_index)
# save to file
summarytable_file2 = here("results", "summary_carbon_index_table.rds")
saveRDS(summary_carbon_index, file = summarytable_file2)
```
```{r}
#carbon_stock data
summary_carbon_stock = skimr::skim(carbon_stock)
print(summary_carbon_stock)
# save to file
summarytable_file3 = here("results", "summary_carbon_stock_table.rds")
saveRDS(summary_carbon_stock, file = summarytable_file3)
```
```{r}
#forest_area data
summary_forest_area = skimr::skim(forest_area)
print(summary_forest_area)
# save to file
summarytable_file3 = here("results", "summary_forest_area_table.rds")
saveRDS(summary_forest_area, file = summarytable_file3)
```
```{r}
#forest_index data
summary_forest_index = skimr::skim(forest_index)
print(summary_forest_index)
# save to file
summarytable_file4 = here("results", "summary_forest_index_table.rds")
saveRDS(summary_forest_index, file = summarytable_file4)
```
```{r}
#forest_share data
summary_forest_share = skimr::skim(forest_share)
print(summary_forest_share)
# save to file
summarytable_file5 = here("results", "summary_forest_share_table.rds")
saveRDS(summary_forest_share, file = summarytable_file5)
```
```{r}
#land_area data
summary_land_area = skimr::skim(land_area)
print(summary_land_area)
# save to file
summarytable_file6 = here("results", "summary_land_area_table.rds")
saveRDS(summary_land_area, file = summarytable_file6)
```


# Data exploration through figures

Now we can explore these data sets through figures. 

It may look confusing or crowded, but let's try getting a line plot showing the climate change over time for each country.

```{r}
temp_organized = subset(temperature, select = -c(ObjectId))
temp_organized <- gather(temp_organized, key = "Year", value = "Temperature Change", 2:63)

# Basic line plot with points
temp_base_plot <- ggplot(data=temp_organized, aes(x=Year, y=`Temperature Change`, group = Country)) +
  geom_line()+
  geom_point() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot(temp_base_plot)

figure_file = here("results","temperature_plot.png")
ggsave(filename = figure_file, plot=temp_base_plot) 

```
It's a bit crowded, but we can see that generally there is an upward trend as the years approach modern day.

Now let's look at what graphs we can make with the Forest and Carbon data.

We will focus on the Forest Area and Carbon Index data sets.



